Exploring the Role of Small Molecules in Biological Systems Using Network Approaches

### 2015-01-19 paper.tex
Updated Role of Networks in Drug Discovery

\section{The Role of Networks in Drug Discovery}
\label{sec:role-networks-drug}
Omics technologies are enabling us today to elucidate and interrogate complex relationships between proteins, genetic material and small molecules at an unprecedented speed and scale. Interrogation of relationships reveal common mechanism of action. This may lead to a therapeutic candidate finding application in a seemingly unrelated disease that share a previously unknown causal mechanism. It also allows identification of drug pairs that show synergistic activity stemming from their interaction with complementary pathways. This type of relationship elucidation may also prevent unwanted side-effects if interrogation would reveal an unexpected link to a vital pathway that ought to remain unperturbed. The process of interrogation starts with experiments aimed at elucidating molecular mechanisms and pathways such as yeast two-hybrid techniques, protein and small molecule mass spectrometry, DNA microarrays and copy number variation studies. Experiments may also include high throughput screening of millions of chemical compounds against specific biological targets of therapeutic interest to find respective agonists and antagonists. These experiments result in pairwise interaction relationships from which the functional big picture biological network can then be elucidated \cite{Wu2010}. Community structure algorithm \cite{girvan2002community} may be used, that allows one to discover a set of sparsely interconnected subsets of vertices, the subsets in turn however being densely connected within themselves. In the biological context, this allows connecting a set of biological cycles and smaller pathways into a big picture biological metacommunity network from which inferences on cross-reactivity and new application areas of a drug candidate can be drawn.

Network based approach to mechanism elucidation of small molecules that allow identification of synergistic pairs have been implemented in tools such as the DrugComboRanker \cite{huang2014drugcomboranker}. The authors use a community based partitioning to create a drug metacommunity network and a disease signaling metacommunity network. Drug targets are then mapped onto disease networks to allow identification of synergistic pairs of drugs and complementary mechanism of action.

>>>>

\section{The Role of Networks in Drug Discovery}
\label{sec:role-networks-drug}
Omics technologies are enabling us today to elucidate and interrogate complex relationships between disease, proteins, genetic material and small molecules at an unprecedented speed and scale. Interrogation of relationships reveal common mechanism of action. This may lead to a therapeutic candidate finding application in a seemingly unrelated disease (repositioning) that share a previously unknown causal mechanism. It also allows identification of drug pairs that show synergistic activity stemming from their interactions with complementary pathways. This type of relationship elucidation may also prevent unwanted side-effects if interrogation would reveal an unexpected link to a vital pathway that ought to remain unperturbed. The process of interrogation starts with experiments aimed at elucidating molecular mechanisms and pathways such as yeast two-hybrid techniques, protein and small molecule mass spectrometry, microarray and copy number variation studies. Experiments may also include high throughput screening of millions of chemical compounds against specific biological targets of therapeutic interest to find respective agonists and antagonists. These experiments result in pairwise interaction relationships from which the functional big picture biological network can then be elucidated \cite{Wu2010}. Algorithms that detect communities from complex networks eg. community structure algorithm \cite{girvan2002community} may be used. These network analyses algorithms allow discovery of a set of sparsely interconnected subsets of vertices where the subsets themselves are however densely connected within. In the biological context, this allows connecting a set of biological cycles and smaller pathways into a big picture biological metacommunity network from which inferences on cross-reactivity and new application areas of a drug candidate can be drawn.

Network based approach to mechanism elucidation of small molecules that allow identification of synergistic pairs have been implemented in tools such as the DrugComboRanker \cite{huang2014drugcomboranker}. The authors use a community based partitioning to create a drug metacommunity network and a disease signaling metacommunity network. Drug targets are then mapped onto disease networks to allow identification of synergistic pairs of drugs and complementary mechanisms of action.
In the tool MANTRA (Mode of Action by NeTwoRk Analysis)\cite{iorio2010discovery}, authors are able to deduce mechanisms of action of new molecules and suggest repositioning of a well known molecule to a new area of application.

Network analysis has also found its application in Medicinal Chemistry and High-throughput Screening. Cheminformatics based chemical scaffold and topology networks capture the structure-activity relationships between compounds in screening experiments \cite{varin2011mining}. These help in medicinal chemistry efforts towards the identification and optimization of lead compounds for the  biological target of interest by allowing visualization of the chemical space in the backdrop of biological activity.
